Confronting preferential sampling in wildlife surveys: diagnosis and model-based triage \footnotemark[2]

### 2016-11-15 manuscript/ReviewerResponses.tex

\documentclass[12pt,fleqn]{article}
%\documentclass[12pt,a4paper]{article}
\usepackage{amsmath,epsfig,epsf,psfrag,color,ecology,natbib}
%\usepackage{a4wide,amsmath,epsfig,epsf,psfrag}
\usepackage[nolists]{endfloat}
\bibliographystyle{ecology}

%  put your commands here

\def\be{{\ensuremath\mathbf{e}}}
\def\bx{{\ensuremath\mathbf{x}}}
\def\bftheta{{\ensuremath\boldsymbol{\theta}}}
\def\bfeta{{\ensuremath\boldsymbol{\eta}}}
\def\bfdelta{{\ensuremath\boldsymbol{\delta}}}
\def\bfSigma{{\ensuremath\boldsymbol{\Sigma}}}
\def\bfB{\textbf{B}}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}

\begin{document}

\bf \hspace{-.32in} Conn et al. responses to reviewer comments

%  make sure that the document has 25 lines per page (it is 12 pt)

\setlength{\textheight}{575pt} \setlength{\baselineskip}{23pt}

%  IMPORTANT -- place the \title, \author, etc, \begin{summary}...
%  \end{summary} \keywords statements in the order demonstrated below
\rm
Here, we provide line by line responses to associate editor and reviewer comments, suggestions, and edits.  For completeness, we include all reviewer comments in addition to our responses (our responses are in blue).  We have also included responses to additional comments we received from a quantitative ecologist (Mark K\'{e}ry) that we hope will improve presentation to non-statisticians.  In addition to these responses, we provide a marked up PDF version of our manuscript showing all changes made.

\section{Associate editor}

[From decision letter]
In view of the comments received, I must decline your manuscript for publication in the Methods in Ecology and Evolution at this time.  However, we would be willing to consider a new manuscript which takes into consideration these comments. In addition to the other comments, I would ask you to try to make the presentation more accessible to an ecological audience - as you put it in the context of SDMs, I think it should really be understandable by ecologists working with SDMs. I appreciate that this can be difficult.

\textcolor{blue}{
We thank the associate editor for the invitation to resubmit, and for their constructive criticism.  In our revision, we have made an effort to make the manuscript more accessible to quantitative ecologists and to those working in the SDM modeling field.  It is unfortunate that there seems to be a different set of traditions (and jargon) between those working in the scientific survey world (e.g., planned surveys of occupancy and abundance), and those working with presence-only data.  Admittedly, there are different challenges in each, but it can make cross-communication somewhat difficult.  In the attached we've tried to emphasize relationships between the two fields and to locate or work within the broader sphere of SDM modeling.}

To give one example, you wrote (p4, l78-80):
``For instance, PS can occur when the investigator uses a priori knowledge or observations of the state variable obtained during sampling to allocate survey effort in places where abundance or occurrence is known to be high."

which could be re-written as

``For instance, PS can occur when the investigator tends to sample more in places where a species has already been observed, or is known to be abundant."

\textcolor{blue} {This formulation is much more to the point; we have accepted this change.}

[From the comments to author section]
Your paper definitely focuses on an important issue, and provides ways to address the problem. Besides the issues raised by the referee, I think you should relate your paper a bit more to the ecological literature, in particular on SDMs (see Edwards et al. 2006). I think in particular that in addition to the types of sampling you discuss, one could also consider sampling designs that are optimal to estimate response curve (eg Albert et al. 2010 - sorry for the self-citation...). Ecological studies may have different objectives, and I think you should try to make your paper relevant to as many readers as possible.

Albert, C. H., N. G. Yoccoz, T. C. Edwards, C. H. Graham, N. E. Zimmermann, and W. Thuiller. 2010. Sampling in ecology and evolution – bridging the gap between theory and practice. Ecography 33:1028-1037.
Edwards, T. C., D. R. Cutler, N. E. Zimmermann, L. Geiser, and G. G. Moisen. 2006. Effects of sample survey design on the accuracy of classification tree models in species distribution models. Ecological Modelling 199:132-141.

\textcolor{blue}{We thank the editor for pointing us to these articles - our knowledge of the species distribution modeling literature is admittedly a little lacking.  These papers are certainly germane to the present study, and we have included additional material in the introduction (2nd paragraph) that talks about (1) implications of design vs. model-based inference, and (2) the importance of the particular objective function (e.g., minimizing variance, estimating species-habitat relationships) for optimal design.}

\section{Reviewer 1}

{\bf A generalized preferential sampling model}

Your random effects model doesn’t appear similar to \citet{RoyleBerliner1999} to me at all, and in fact this suggestion led to some confusion on my part about how these random effects were specified. The joint
specification of $\bfeta$ and $\bfdelta$, following \citet{RoyleBerliner1999} and assuming 0 mean, is:
\begin{equation}
  \left(
    \begin{array}{c}
    \bfeta \\
    \bfdelta
    \end{array}
   \right)
   \sim \textrm{Gaussian}
   \left(
     \left(
       \begin{array}{c}
       0 \\
       0
       \end{array}
     \right),
     \left(
       \begin{array}{cc}
         \bfSigma_\eta + \bfB \bfSigma_\delta \bfB^\prime & \bfB \bfSigma_\delta \\
         (\bfB \bfSigma_\delta)^\prime & \bfSigma_\delta
       \end{array}
     \right)
   \right),
\end{equation}
which has a very different covariance structure than the random effects model presented in Appendix S1.

\textcolor{blue}{There are several points to make here.  First, we never claimed that we were fitting the exact model proposed by \citet{RoyleBerliner1999} - we're not.  However, the covariance structure implied by the two model \emph{is} actually very similar.  For instance, if $\bfeta \sim \text{Gaussian}({\bf 0},\bfSigma_\eta)$ and $\bfdelta \sim \text{Gaussian}({\bf 0},\bfSigma_\delta)$, then
\begin{equation}
  \text{Cov}\left(
    \begin{array}{c}
    \bfeta + \bfB \bfdelta \\
    \bfdelta
    \end{array}
   \right)
   =
   \left(
   \begin{array}{cc}
      \bfSigma_\eta + \bfB \bfSigma_\delta \bfB^\prime & \bfB \bfSigma_\delta \\
         (\bfB \bfSigma_\delta)^\prime & \bfSigma_\delta
    \end{array}
   \right),
\end{equation}
as desired.  Second, it seems less important that the model be exactly the same as it does to provide a motivation (and citation) for how the $\bfB$ matrix can be specified.  We think both of these points were adequately accounted for in our initial submission, as we (a) described the \citet{RoyleBerliner1999} model as being ``similar" to the one we proposed here, and (b) used their paper as a reference for how the $\bfB$ matrix can be formulated and interpreted.
}

In your application, line 199, you report difficulties fitting a model with even modest structure on $\bfB$. Might a
joint model for $\bfeta$ and $\bfdelta$, as specified by \citet{RoyleBerliner1999}, allow you to specify a more complicated
model for $\bfB$?

\textcolor{blue}{
  As suggested above, we believe our model to be very similar to that proposed by \citet{RoyleBerliner1999}.  We're not seeing how drawing the random effects jointly would necessarily improve performance (we're also having trouble seeing how we could implement this in our chosen software).  Also, after further investigation of the \citet{RoyleBerliner1999} paper, we're not sure they ever actually fit a model with a more complicated relationship in $\bfB$.  In fact, their one example on ozone-temperature analysis uses the same simple diagonal parameterization we used here.  So we think the question remains whether more complicated models for $\bfB$ are ever identifiable, and if so, whether they are reasonable to fit with data sets typical of ecological applications.  We've adjusted text in several places to reflect this. For instance, we state that ``However, to our knowledge the identifiability of higher order models (such as trend surfaces) has not been investigated" when talking about higher order preferential sampling models.  In the discussion, we indicate that ``We attempted to fit models to bearded seal data where the degree of PS changed over the landscape, as \citet{RoyleBerliner1999} suggested might be possible (but never demonstrated) for multivariate spatial models."  We also suggest that identifiability of such models be further investigated in future research.
}

On line 139, you suggest “we often need to fix $\beta_0^* = 0.0$ . . . to permit parameter identification”. The word
``often" suggests you can sometimes include the intercept parameter when estimating random effects. Can
you provide more guidance about when you would need to fix the intercept at 0, and how a practitioner
would know they need to fit the intercept at 0?

\textcolor{blue}{
  This wording was too ambiguous, and was likely a result of initially fitting different types of spatial models to the data.  For instance, ICAR type models have a known parameter redundancy built in, such that the intercept is confounded with the spatial process \citep{RueHeld2005}.  The type of Gaussian process model we settled on and used in analysis does not suffer from this deficiency.  Further confusion resulted from incorrectly specifying a multinomial logit link function in our initial analyses.  We have rectified this mistake in the present version, and all models for $R_i$ are now Bernoulli with a logit link, so that the intercept term is now identifiable.  We now report results for this corrected analysis.
}

{\bf Notation}

I have difficulty following your notation at several points throughout the manuscript, which are outlined
below.

On line 41, you state $Z_i$ is the state variable of interest (e.g., occupancy or abundance), which I take will
be the notation you use throughout given the section header. Your specication for a count model on eqn
4 therefore becomes confusing. In particular, if $Z_i$ represents abundance, for reasonable values of $A_i$ and $p_i$ you could have random realizations $Y_i$ from a Poisson distribution that are greater than abundance, $Z_i$. If $Z_i$ does not represent abundance, what state process is represented here?

\textcolor{blue}{
  We have replaced the Poisson observation model with a binomial one in eqn 4 to make this clearer, and have indicated dependence on $Z_i$ through conditioning.  The reason that we had the Poisson in there originally is that the models we fit to data use the fact that the marginal distribution for $Y_i$ given the Poisson-Binomial mixture is itself a Poisson distribution.  However, we think it will be clearer to present the model hierarchically at this stage and then to describe our marginalization (something many ecologists may not care about) when describing software.  We have created a new ``Software" methods subsection which describes this in greater detail. Importantly, we describe how predictions are made; when detection probability is equal to 1.0, predictions are made using the formulation $\hat{N}_i = Y_i + (1-A_i)\exp(\hat{\mu}_i)$.  This framework incorporates a finite population correction and is similar to that employed by \citet{JohnsonEtAl2010} in the context of point process likelihood analysis of distance sampling data.  It also prevents predicted abundance in a cell from being less than the observed count.
}

\section{Additional reviewer: Marc K\'{e}ry}

I read your ms and found it very stimulating and also well-written. Just a couple of minor comments which I noted during my reading.
-          P.3, eq. 4: why do you put a Poisson observation model here ? In particular, since the Poisson has no “ceiling”, p cannot be detection probability but must be a detection rate, i.e., the combination of false-positive and false-negative errors. I note that in the Appendix 1 you have the usual Binomial observation model

\textcolor{blue}{This was also noted by Reviewer 1. There are two alternatives here.  In the first, one can use some sort of simulation-based inference (e.g. MCMC) to model both components of the hierarchy (i.e. $[{\bf Z}|\bftheta]$ and $[{\bf Y}|{\bf Z}]$), in which case one could specify $[{\bf Z}|\bftheta]$ to be Poisson and $[{\bf Y}|{\bf Z},p]$ to be binomial.  Keeping track of the latent $Z_i$s, one then has an entirely coherent model for abundance.  However, from a computational standpoint, it is much easier to deal with the unconditional distribution $[{\bf Y}|\bftheta,p]$.  If $\bftheta$ are the expectations of ${\bf Z}$, then unconditionally $[{\bf Y}|\bftheta,p] = \text{Poisson}(\bftheta p)$ (apologies for a little sloppiness in notation here).   From a computational standpoint, the Poisson distribution is much easier to deal with, as one can bring the TMB machinery into the analysis (in particular, since it works with derivatives, it can't handle discrete random effects). The inference about the parameters $\bftheta$ should be completely coherent. The disadvantage is that one must then predict the latent ${\bf Z}$ parameters: e.g. once we have conducted inference relative to $\bftheta$, one must then predict $Z_i \sim \text{Poisson}(\theta_i)$. This isn't perfect, as we could potentially have the case that the predicted $Z_i$ is actually less that the count $Y_i$.  However, it should be a reasonable approximation, especially when focus is on absolute abundance (i.e. $\sum Y_i$) and will actually be a good approximation at the individual cell level whenever $Z_i$ is large and $p$ is small (e.g., when only a small fraction of each survey unit is sampled, as in the bearded seal example).  We note that a) this strategy of inference has been used in several recent applications with good results (e.g. \cite{ThorsonEtAl}, and b) the simulation results suggest this procedure is unbiased (see Fig. 4, Independent estimation model, $b=0$ case).  Note that in simulations, we did generate data as $[{\bf Z}] = \text{Poisson}$ and $[{\bf Y} | {\bf Z}] = \text{Binomial}$.  We now discuss this tradeoff more explicitly, both in Appendix S1 and in the Discussion.
}

Line 68: What are platforms of opportunity ? perhaps sea mammal jargon ? Hard to understand for non-sea-mammal guys like myself

\textcolor{blue}{This is a good point - it is actually marine mammal jargon (for example, U.S. military ships providing locations where whale observations are made opportunistically).  We deleted this term, replacing it with ``opportunistic sampling."}

-          Line 139: doesn’t setting beta-star-0 to zero imply that the probability of a sample unit to be surveyed is fixed at 1 for the average environmental conditions (if covariates are centred) ?
-          168: lest
-          199: I would say “and estimates are not reported here”
-          209 and 211: fig. instead of Fig
-          317: typo
-          Fig. 1: main title left: Coarse scale
-          Fig. 4, legend: since you’re doing ML you may briefly comment on why you have a posterior mode here ? Presumably some application of Bayes rule for empirical Bayes estimates ?

\bibliography{master_bib}

\end{document}

### 2016-11-15 manuscript/pref_sampling.tex
Commit during ms. revisions.  Changed link function of Bernoulli sampling model to be logit and now include an intercept.  Simulation loop fails (crashes R) in TMB_version/examples/simple_example_v6.R

\def\volumeyear{2015}
\def\VOC{VO}

@@

\item Using simulation, we show that moderate levels of preferential sampling can lead to large (e.g. $40\%$) bias in estimates of animal density, and that our modeling approach can considerably reduce bias.

@@

\item When wildlife surveys are conducted without a well defined sampling frame, ecologists should recognize the potentially biasing effects of preferential sampling. Joint models, such as those described in this paper, can be used to test and correct for such biases.  Predictive covariates are also useful for bias reduction, but ultimately the best way to avoid preferential sampling bias is to incorporate design-based principles such as randomization and/or systematic sampling into survey design.
\end{enumerate}

@@

Surveys of unmarked animal populations are often used to estimate abundance and occurrence of animal populations and to predict species distributions, enterprises central to conservation, ecology, and management. For studies of abundance, researchers historically relied on design-based statistical inference \citep[e.g.][]{Cochran1977}, which requires adoption of a pre-defined sampling frame (e.g. using systematic random sampling, stratified random sampling, or some variant thereof).  Designing animal surveys is relatively straightforward in such applications, and unbiased point and variance estimators are available.  Recently, however, there has been a surge in research describing model-based procedures for estimating abundance, density, and occupancy from surveys of unmarked animals, including N-mixture and Dail-Madsen models for repeated point counts \citep{Royle2004a,DailMadsen2011}, occupancy models for presence-absence surveys \citep{MacKenzie2002,JohnsonEtAl2013}, and various model-based formulations for distance-sampling data \citep{HedleyBuckland2004,MillerEtAl2013,JohnsonEtAl2010}.  In such applications, it is common to use habitat or environmental covariates together with spatial effects (e.g. via trend surfaces or spatial random effects) to predict density or distributions across the landscape.  We shall refer to the amalgam of model-based approaches for making spatially explicit inference about animal populations as ``Species distribution models" \citep[SDMs; {\it sensu}][]{ElithLeathwick2009}, even though this term is more often used to refer to animal occurrence than it is to density or abundance.

One of the main advantages of using SDMs is that one is no longer beholden to predetermined sampling frames, and can potentially use data gathered from non-randomized designs or platforms of opportunity to make inferences about animal populations \citep{JohnsonEtAl2010}.  However,
in a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling disproportionately targets locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.  In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance, regardless of predictive covariates. For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.  Bias might also arise if surveying effort is higher near bases of operations, and if animal abundance is higher (or lower) near bases of operations than elsewhere in the landscape.

@@

However, the manner in which effort is ultimately allocated can potentially have profound influence on SDM estimator performance.  With respect to nonrandom sampling, two possible problems seem particularly likely in discrete spatial domains: coarse scale preferential sampling (CSPS), and fine scale preferential sampling (FSPS) (Fig. \ref{fig:pref}).  FSPS arises when the observations taken at a  particular sampling unit are non-random with respect to the density of animals within that sampling unit.  For instance, when allocating line transect survey effort, it may be tempting to place the transect in a manner that targets habitat or landscape features that maximize the number of animals that will be encountered.  Depending upon the interpretation of occupancy, this may or may not be reasonable.  However, if trying to estimate density or abundance, this strategy will clearly lead to positive bias.

@@

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurence) and the locations chosen for sampling. For the abundance process model, we start with eq. \ref{eq:process} as a general formulation for non-Gaussian data, writing the link-scale expectation as in eq. \ref{eq:mu}.
Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is selected for sampling, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
\begin{eqnarray}
 \label{eq:R}
  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
\begin{eqnarray}
  \label{eq:nu}
  \boldsymbol{\nu} & = & \beta_0^* + {\bf x}^* \boldsymbol{\beta}^*+\boldsymbol{\eta}+{\bf B} \boldsymbol{\delta}.
\label{eq:nu}
\end{eqnarray}
 In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$) and spatially autocorrelated random effects ($\boldsymbol{\eta}$ and $\boldsymbol{\delta}$).  The predictive covariates ${\bf x}_i$ from Eq. \ref{eq:mu} and ${\bf x}_i^*$ from Eq. \ref{eq:nu} may or may not be the same.  Note also that the  spatially autocorrelated random effects $\boldsymbol{\delta}$ are included in both Eqs. \ref{eq:mu} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.  The spatially autocorrelated random effects $\bfeta$ are assumed independent of the $\bfdelta$.  In practice, we find we often need to fix $\beta_0^*=0.0$ when random effects in Eq. \ref{eq:nu} are estimated to permit parameter identification.

@@

We used maximum likelihood to conduct statistical inference. In particular, we used Template Model Builder \citep[TMB; ][]{KristensenEtAl2015}, interfaced with the R programming environment, to conduct maximization. The TMB software uses a Laplace approximation to integrate out random effects ($\bfeta$ and $\bfdelta$), and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.  This approach resulted in a facile implementation and speedy computing times, allowing us to conduct simulation and model testing with greater efficiency than would have been possible with Bayesian simulation.  Further detail on statistical methods are provided in Appendix S1; requisite R and TMB code will be published to a publicly accessible repository upon acceptance, and is also available at \url{https://github.com/NMML/pref_sampling/}.

@@

In this study, we showed that course-grained preferential sampling (Fig. \ref{fig:pref}) can have a profound impact on the quality of estimates (e.g. animal abundance) when sampling is non-randomized. In simulations, estimators were increasingly positively biased as PS increased. When PS was present, we were able to substantially reduce bias by conducting estimation under a framework where the state variable of interest and the sites chosen for sampling were jointly modeled under a dependent covariance structure.  In absence of PS, simulations suggest that this structure results in lower precision than a model without a PS effect; thus the need to account for PS reduces the quality of inference.

@@

Model-based approaches to estimation of abundance or occurrence have become popular in recent years.  We (the authors) have noticed a tendency for analysts to assume that inclusion of spatial covariates or random effects into predictive models will make the underlying sampling design ignorable. We have shown in this paper that this is not the case, although our results do suggest that including predictive covariates can indeed decrease bias from preferential sampling.  We have also shown that it is possible to further diagnose and adjust for preferential sampling by jointly modeling dependence between the data collection mechanism and the process of interest (e.g. abundance or occupancy).  However, such models can be considerably less precise and have greater instability than models without a preferential sampling parameter and it would be preferable not to have to go to such elaborations. Where possible, we suggest that survey planners incorporate design-based elements (e.g. random or systematic sampling) into their survey designs to reduce the need for model-based triage.

>>>>

\def\volumeyear{2016}
\def\VOC{VO}

@@

\item Using simulation, we show that moderate levels of preferential sampling can lead to large (e.g. $40\%$) bias in estimates of animal density, and that our modeling approach can considerably reduce this bias.

@@

\item When wildlife surveys are conducted without a well-defined sampling frame, ecologists should recognize the potentially biasing effects of preferential sampling. Joint models, such as those described in this paper, can be used to test and correct for such biases.  Predictive covariates are also useful for bias reduction, but ultimately the best way to avoid preferential sampling bias is to incorporate design-based principles such as randomization and/or systematic sampling into survey design.
\end{enumerate}

@@

Surveys of unmarked animal populations are often used to estimate abundance and occurrence of animal populations and to predict species distributions, enterprises central to conservation, ecology, and management. For studies of abundance, researchers historically relied on design-based statistical inference \citep[e.g.][]{Cochran1977}, which requires adoption of a pre-defined sampling frame (e.g. using systematic random sampling, stratified random sampling, or some variant thereof).  Designing animal surveys is relatively straightforward in such applications, and unbiased point and variance estimators are available.  Recently, however, there has been a surge in research describing model-based procedures for estimating abundance, density, and occupancy from surveys of unmarked animals, including N-mixture and Dail-Madsen models for repeated point counts \citep{Royle2004a,DailMadsen2011}, occupancy models for presence-absence surveys \citep{MacKenzie2002,JohnsonEtAl2013}, and various model-based formulations for distance-sampling data \citep{HedleyBuckland2004,JohnsonEtAl2010,MillerEtAl2013}.  In such applications, it is common to use habitat or environmental covariates together with spatial effects (e.g. via trend surfaces or spatial random effects) to predict density or distributions across the landscape.  We shall refer to the amalgam of model-based approaches for making spatially explicit inference about animal populations as ``species distribution models" \citep[SDMs; {\it sensu}][]{ElithLeathwick2009}, even though this term is more often used to refer to animal occurrence than it is to density or abundance.

One of the main advantages of using SDMs is that one is no longer beholden to predetermined sampling frames, and can potentially use data gathered from non-randomized designs or platforms of opportunity to make inferences about animal populations \citep{JohnsonEtAl2010}.  However,
in a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling disproportionately targets locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.  In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance. For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.  Bias might also arise if surveying effort is higher near bases of operations, and if animal abundance is higher (or lower) near bases of operations than elsewhere in the landscape.

@@

However, the manner in which effort is ultimately allocated can potentially have profound influence on SDM estimator performance.  With respect to nonrandom sampling, two possible problems seem particularly likely in discrete spatial domains: coarse scale preferential sampling (CSPS), and fine scale preferential sampling (FSPS) (Fig. \ref{fig:pref}).  FSPS arises when the observations taken at a  particular sampling unit are non-random with respect to the density of animals within that sampling unit.  For instance, when allocating line transect survey effort, it may be tempting to place the transect in a manner that targets habitat or landscape features that maximize the number of animals that will be encountered.  However, this strategy will clearly lead to positive bias when estimating density or abundance.

@@

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurrence) and the locations chosen for sampling. For the abundance process model, we start with eq. \ref{eq:process} as a general formulation for non-Gaussian data, writing the link-scale expectation as in eq. \ref{eq:mu}.
Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is selected for sampling, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
\begin{eqnarray}
 \label{eq:R}
  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
\begin{eqnarray}
  \label{eq:nu}
  \boldsymbol{\nu} & = & \beta_0^* + {\bf x}^* \boldsymbol{\beta}^*+\boldsymbol{\eta}+{\bf B} \boldsymbol{\delta}.
\label{eq:nu}
\end{eqnarray}
 In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$) and spatially autocorrelated random effects ($\boldsymbol{\eta}$ and $\boldsymbol{\delta}$).  The predictive covariates ${\bf x}_i$ from Eq. \ref{eq:mu} and ${\bf x}_i^*$ from Eq. \ref{eq:nu} may or may not be the same.  Note also that the  spatially autocorrelated random effects $\boldsymbol{\delta}$ are included in both Eqs. \ref{eq:mu} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.  The spatially autocorrelated random effects $\bfeta$ are assumed independent of the $\bfdelta$.  In practice, we find we often need to fix $\beta_0^*=0.0$ when random effects in Eq. \ref{eq:nu} are estimated to permit parameter identification.

@@

We used maximum likelihood to conduct statistical inference. In particular, we used Template Model Builder \citep[TMB; ][]{KristensenEtAl2016}, interfaced with the R programming environment, to conduct maximization. The TMB software uses a Laplace approximation to integrate out random effects ($\bfeta$ and $\bfdelta$), and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.  This approach resulted in a facile implementation and speedy computing times, allowing us to conduct simulation and model testing with greater efficiency than would have been possible with Bayesian simulation.  Further detail on statistical methods are provided in Appendix S1; requisite R and TMB code will be published to a publicly accessible repository upon acceptance, and is also available at \url{https://github.com/NMML/pref_sampling/}.

@@

In this study, we showed that coarse-grained preferential sampling (Fig. \ref{fig:pref}) can have a profound impact on the quality of estimates (e.g. animal abundance) when sampling is non-randomized. In simulations, estimators were increasingly positively biased as PS increased. When PS was present, we were able to substantially reduce bias by conducting estimation under a framework where the state variable of interest and the sites chosen for sampling were jointly modeled under a dependent covariance structure.  In absence of PS, simulations suggest that this structure results in lower precision than a model without a PS effect; thus the need to account for PS reduces the quality of inference.

@@

Model-based approaches to estimation of abundance or occurrence have become popular in recent years.  We (the authors) have noticed a tendency for analysts to assume that inclusion of spatial covariates or random effects into predictive models will make the underlying sampling design ignorable. We have shown in this paper that this is not the case, although our results do suggest that including predictive covariates can indeed decrease bias from preferential sampling.  We have also shown that it is possible to further diagnose and adjust for preferential sampling by jointly modeling dependence between the data collection mechanism and the process of interest (e.g. abundance or occupancy).  However, such models can be considerably less precise and have greater instability than models without a preferential sampling parameter. Where possible, we suggest that survey planners incorporate design-based elements (e.g. random or systematic sampling) into their survey designs to reduce the need for model-based triage.
